```{r}
#| echo: false
source("code/before_script.R")
```

# **tmap** in a nutshell {#sec-nutshell}

The **tmap** package allows the creation of thematic maps with great flexibility.
It accepts spatial data in various formats -- shape objects (@sec-shape-objects)
Next, the data can be used to create simple, quick maps (@sec-quick-maps) and more complex and expandable maps (@sec-regular-maps).
These maps can be presented in two modes: as a static map and an interactive one.
Additionally, **tmap** makes it possible to create facet maps (@sec-facets-section) and map animations (@sec-ani-section).
The goal of this chapter is to provide a brief overview of the main **tmap** features.

## Shape objects {#sec-shape-objects}

As we established in @sec-geodata, spatial data comes in various file formats related to two main data models -- vector and raster.
There are also several spatial object classes in R, for example, `sf` from the **sf** package for vector data and `stars` from **stars** for raster data and spatial data cubes.
Additionally, packages such as **sp**, **raster**, or **terra** offer their own classes, and this abundance of spatial object classes can be generally overwhelming.
Gladly, **tmap** can work with all of the above objects -- it treats all supported spatial data classes as so-called *shape objects*.

For example, we read the `slo_cities.gpkg` file containing several point locations representing cities and towns in Slovenia into a new object, `slo_cities`.
The `slo_cities` object is a *shape object*.

```{r}
#| message: false
library(tmap)
library(sf)
slo_cities = read_sf("data/slovenia/slo_cities.gpkg")
```

<!-- ref also to data appendix -->

Spatial data, no matter the class, usually stores two interrelated sets of information - about the locations/geometries and their associated values/attributes.
Visualization of the attributes only can be done with regular plotting functions (e.g., `plot()`, `hist()`, `barplot()`) or dedicated packages, such as **ggplot2** <!--cite-->.
On the other hand, **tmap** is suitable when our goal is to visualize spatial geometries only or spatial geometries together with their attributes.

## Quick maps {#sec-quick-maps}

The **tmap** package offers a distinction between quick and regular maps.
The first approach, using the `qtm()` function, could be handy for data exploration.
It works even if we just provide any *shape object* -- in that case, only the geometry is plotted.
@fig-qtm-1 shows a visualization of the geometries from the `slo_cities`.

```{r}
#| label: qtm1
#| eval: false
qtm(slo_cities)
```

The `qtm()` function allows to customize many map elements for the provided *shape object*.
For example, we can change the shapes of the points in `slo_cities`, make their sizes related to the the `"population"` argument, and add a title (@fig-qtm-2).

```{r}
#| label: qtm2
#| eval: false
qtm(slo_cities, shape = 24, size = "population", title = "Cities")
```

```{r}
#| label: fig-qtm
#| echo: false
#| layout-ncol: 2
#| fig-cap: Example of a quick map created with the `qtm()` function.
#| fig-subcap: 
#|   - A map with geometries only.
#|   - A map with geometries and attributes.
<<qtm1>>
<<qtm2>>
```

<!-- The `qtm()` function offers similar flexibility to the regular map approach.  -->
<!-- It only supports one shape object at a time, but many `qtm()` calls can be combined in one code chunk. -->

## Regular maps {#sec-regular-maps}

Therefore, for most applications, we recommend using the regular mapping approach.
This approach operates on many functions that start with `tm_`. 
The first element always^[Almost always...] is `tm_shape()`, which specifies the input shape object.
Next, map layers, additional map elements, and overall layout options can be customized.
<!-- update terminology later -->
<!-- references -->

The last example in @sec-quick-maps can be reproduced with the regular map approach using the following code.

```{r}
#| eval: false
tm_shape(slo_cities) +
  tm_symbols(shape = 24, size = "population") +
  tm_title("Cities")
```

Here, we specify the input data (our shape object) with `tm_shape()`, aesthetics (also known as *visual variables*) of map layers with `tm_symbols()`, and the map title with `tm_title()`.

The **tmap** package has a number of possible map layers, but the most prominent ones are `tm_polygons()`, `tm_symbols()`, `tm_lines()`, `tm_raster()`, and `tm_text()` (@sec-layers).
Overall, most visual variables of map layers can be assigned in two main ways.
First, they accept a fixed, constant value, for instance, `shape = 24`, which sets the symbols' shapes to triangles.
Second, it is also possible to provide a variable name, for example `size = "population"`.
This plots each point with a size based on the `population` attribute from the `slo_cities` object and automatically adds a related map legend.

The `tm_shape()` function and one or more following map layers create a *group* together.
In other words, map layers are related only to the preceding `tm_shape()` call.
One map can have several *groups*.
Let's see how many *groups* can be used by reading some additional datasets -- the `slo_elev` raster with elevation data of Slovenia, the `slo_borders` polygon with the country borders, and the `slo_railroads` lines contains a railroad network for this country.
<!-- ref data appendix -->

```{r}
#| message: false
library(sf)
library(stars)
slo_elev = read_stars("data/slovenia/slo_elev.tif")
slo_borders = read_sf("data/slovenia/slo_border.gpkg")
slo_railroads = read_sf("data/slovenia/slo_railroads.gpkg")
```

Look at the following example and try to guess how many *groups* it has, and how many layers exist for each *group* (@fig-rmap1).

```{r}
#| label: fig-rmap1
#| warning: false
#| fig-cap: "Example of a map with four groups of map layers: an elevation layer, country borders layer, railroads layer, and cities layer."
tm_shape(slo_elev) +
  tm_raster(col.scale = tm_scale(values = "geyser"),
            col.legend = tm_legend(title = "Elevation (m asl)")) +
  tm_shape(slo_borders) + 
  tm_borders() +
  tm_shape(slo_railroads) +
  tm_lines(lwd = "track_width", 
           lwd.legend = tm_legend(show = FALSE)) +
  tm_shape(slo_cities) +
  tm_symbols(shape = 24, size = "population",
             size.legend = tm_legend(title = "Population")) +
  tm_title("Slovenia") +
  tm_layout(bg.color = "grey95")
```

The correct answer is four groups, all with just one layer.
Each *group* is put on top of the previous one -- **tmap** uses a layered approach.
The first *group* represents elevation data with a continuous color scale style, a color palette called `"geyser"`, and a legend title.
The second *group* shows the borders of Slovenia with the default aesthetics, while the third *group* presents the railroad network (the `slo_railroads` object), with each line's width based on the values from the `"track_width"` column, but with a legend hidden.
The last *group* is similar to our previous example with fixed symbol shapes and sizes related to the `"elevation"` attribute, but also with the legend title instead of the map title.
Additionally, we use the `tm_title()` function to add a map title and `tm_layout` to modify the general appearance of the map.
You can also notice that we can control scales of various visual variables, such as color, size, or width, with the `tm_scale_*()` function and customize legends with the `tm_legend()` function.
<!-- ref to other parts of the book -->

Often, maps also have additional map elements, such as graticule lines, north arrow, scale bar, or map credits (@fig-rmap2). 
They help map readers understand the location or extent of the input data and provide some ancillary information.
The **tmap** package offers a set of functions for additional map elements.
The `tm_graticules()` function draws latitude and longitude graticules and adds their labels.
It also uses the layered approach, and thus, the lines will be drawn either below or above the shape objects, depending on the position of this function in the code.
In our example below, `tm_graticules()` is used after all of the map groups, and that is why the graticule lines are put on the top of the spatial data.
We can also use `tm_compass()` to create a north arrow, `tm_scalebar()` to add a scale bar, and `tm_credits()` to add a text annotation representing credits or acknowledgments.
The location of all these three elements on the map is, by default, automatically determined.
It, however, can be adjusted with the `position` argument -- see an example of its use in the `tm_compass()` function below.
Moreover, it is possible to add any type of manual legend with `tm_add_legend()`.
It includes simple legends below, such as the `"Railroads"` legend element, that is only represented by a single black line and a related label, but more complex custom legends with several elements are also possible. 
<!-- ref to other parts of the book -->

```{r}
#| warning: false
my_map = tm_shape(slo_elev) +
  tm_raster(col.scale = tm_scale(values = "geyser"),
            col.legend = tm_legend(title = "Elevation (m asl)")) +
  tm_shape(slo_borders) + 
  tm_borders() +
  tm_shape(slo_railroads) +
  tm_lines(lwd = "track_width", 
           lwd.legend = tm_legend(show = FALSE)) +
  tm_shape(slo_cities) +
  tm_symbols(shape = 24, size = "population",
             size.legend = tm_legend(title = "Population")) +
  tm_graticules() +
  tm_compass(position = c("right", "top")) +
  tm_scalebar() +
  tm_credits("Author, Year") +
  tm_add_legend(type = "lines", col = "black", labels = "Railroads") +
  tm_title("Slovenia") +
  tm_layout(bg.color = "grey95")
```

Maps created with **tmap** can be saved as an R object.
This is a useful feature that allows to use one map in a few places in a code, modify existing **tmap** objects, or save these objects to files.

```{r}
#| label: fig-rmap2
#| message: false
#| fig-cap: Example of a map with four groups of map layers and additional map elements, such as graticule lines, north arrow, scale bar, and text annotation. It also has a manually added legend.
my_map
```

<!-- refs  -->

## Facets {#sec-facets-section}

\index{facets}
\index{small multiples}
The **tmap** package allows the creation of facet maps, also known as small multiples.
In such maps, the same data is visualized in several panels, each showing a different subset of the data.
The panels may represent different variables, time periods, or spatial extents.

The `slo_regions` object contains the administrative regions of Slovenia with their names, population density (`gdppercap`), GDP per capita (`gdppercap`), region group name (`region_group`), and other attributes.
One way to visualize this data as facets is to specify selected columns as visual variables in the map layer function (e.g., `tm_polygons()`) (@fig-sec-facets1).

```{r}
#| label: fig-sec-facets1
#| fig-cap: A facet map with three variables visualized in separate panels.
#| fig-asp: 0.5
slo_regions = read_sf("data/slovenia/slo_regions.gpkg")
tm_shape(slo_regions) +
  tm_polygons(fill = c("pop_dens", "gdppercap", "region_group"))
```

Another possible approach is to use the `tm_facets()` function.
Let's illustrate it with the `slo_regions_ts` object containing the same variables as `slo_regions` but for many years (`time` column).
@fig-sec-facets2 shows a facet map with the GDP per capita, specified with `tm_polygons()`, and visualized in separate panels for each year as defined by the `tm_facets()` function.

```{r}
#| label: fig-sec-facets2
#| fig-cap: A facet map with the GDP per capita visualized in separate panels for each year.
slo_regions_ts = read_sf("data/slovenia/slo_regions_ts.gpkg")
tm_shape(slo_regions_ts) +
  tm_polygons("gdppercap") +
  tm_facets(by = "time", ncol = 5)
```

Facet maps can be used for vector and raster data, their extents can be adjusted, and they can be further customized with additional labels.
This is a topic of the @sec-facets.

## Animations {#sec-ani-section}

<!-- to write after the \@ref(animations) chapter is done -->

@sec-animations

## Map modes {#sec-map-modes}

Each map created with **tmap** can be viewed in one of two modes: `"plot"` and `"view"`.
<!-- maybe add note that more modes are possible + ref -->
The `"plot"` mode is used by default and creates static maps similar to those shown before in this chapter.
This mode supports almost all of **tmap**'s features, and it is recommended, for example, for scientific publications or printing.

The second mode, `"view"`, allows the creation of interactive maps.
They can be zoomed in and out or panned, allow for changing background tiles (*basemaps*), or click on map objects to get some additional information.
This mode has, however, some constraints and limitations comparing to `"plot"`, for example, the legend cannot be fully customized, and some additional map elements are not supported.

Both modes can be used on the same **tmap** code.
Therefore, there is no need to create two separate maps for static and interactive use.
The `tmap_mode()` function can be used to switch from one mode to the other^[Map modes can be also changed globally using `tmap_options()` or switched using `ttm()`.].

```{r}
tmap_mode("view")
```

The above line of code just changes the mode -- it does not return anything except a message.
Now, if we want to use this mode, we need to either write a new **tmap** code or provide some existing **tmap** object, such as `my_map`.

```{r}
#| eval: false
my_map
```

Our main result is the interactive map (@fig-imap1).
<!-- improve -->
It shows our spatial data using aesthetics similar to @fig-rmap2 but allows us to zoom in and out or move the map.
We also can select a basemap or click on any line and point to get some information.

```{r}
#| label: fig-imap1
#| echo: false
#| cache: false
#| message: false
#| fig-cap: Map from the previous figure shown using the interactive ("view") mode.
view_map(my_map, "imap1")
```

To go back to the `"plot"` mode, we need to use the `tmap_mode()` function again -- map not shown:

```{r}
#| fig-show: hide
#| message: false
tmap_mode("plot")
my_map
```

More information about the interactive `"view"` mode and how to customize its outputs is in @sec-interactive.
